---
title: "Investigating Migration Flows in and out of NYS"



---

```{r load libraries, message = FALSE}
library(tidyverse)
library(readxl)
library(here)
library(gt)
```

# Read in Data

## State to state migration Data
```{r message = FALSE, warning = FALSE}
ss_23 <- read_xlsx(here("data", "state_to_state_migration/2023.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         ) %>% 
  #slice(-1)

# I had to clean this data before I imported it because it had lots of messy tabls that did not download well
ss_22 <- read_xlsx(here("data", "state_to_state_migration/2022.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )

# I had to clean this data before I imported it because it had lots of messy tabls that did not download well
ss_21 <- read_xlsx(here("data", "state_to_state_migration/2021.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )
ss_19 <- read_xlsx(here("data", "state_to_state_migration/2019.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )


## to remove rows that contain all NAs
#filter(rowSums(is.na(ss_22)) != ncol(ss_22))
```

Only does 2015-022
```{r}
# Directory where the Excel files are stored
file_dir <- here("data", "state_to_state_migration")

# List all Excel files in the directory
files <- list.files(file_dir, pattern = "*.xlsx", full.names = TRUE)

# Define a function to process each file
process_file <- function(file_path) {
  # Extract year from the file name (assuming year is in the filename)
  year <- sub(".*([0-9]{4}).xlsx", "\\1", basename(file_path))
  
  # Read and process the data
  df <- read_xlsx(file_path, skip = 4) %>%
    rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )  # different state as 1 year ago
    
  
  # Create the column names from the state row
  for (i in seq(10, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(df[1, i])  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the state row
  for (i in seq(3, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(df[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  df <- df %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>% 
    slice(-c(1:2))  %>% 
    rename(tot_abroad = Total) %>% 
    mutate_at(vars(-1), as.numeric)
  
  
  # Dynamically assign the processed dataframe to a variable named sb_<year>
  assign(paste0("migration_", year), df, envir = .GlobalEnv)
}

# Apply the function to all files in the folder
lapply(files, process_file)
```

Something was weird for 2023 so I had to do it separately:
```{r}
# Read and process the data
  df <- read_xlsx(here("data", "state_to_state_migration", "2023.xlsx"), skip = 4) %>%
    rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         ) %>%  # different state as 1 year ago
  slice(-1)
    
  
  # Create the column names from the state row
  for (i in seq(10, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(df[1, i])  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the state row
  for (i in seq(3, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(df[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  df <- df %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>%
    slice(-c(1:2))  %>%
    rename(tot_abroad = Total) %>% 
    mutate_at(vars(-1), as.numeric)
  
    
  clean_df[[8]] <- df
  
  # Dynamically assign the processed dataframe to a variable named sb_<year>
  assign(paste0("migration_", 2023), df, envir = .GlobalEnv)
```

## State to state migration

2023:
Net migration 
Into NY from other states
Into NY from Abroad

### Clean data
```{r message = FALSE, warning = FALSE}
list_df <- list(ss_23, ss_22, ss_21, ss_19)
clean_df <- list()

for (a in 1:length(clean_df)) {
  
  d <- list_df[[a]]
  
  # Create the column names from the state row
  for (i in seq(10, ncol(d), by = 2)) {
    if (i <= ncol(d)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(d[1, i])  # Get the value from row 1 of the ith column
      colnames(d)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the state row
  for (i in seq(3, ncol(d), by = 2)) {
    if (i <= ncol(d)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(d[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(d)[i] <- col_name  # Set column name to that value
    }
  }
  
  d <- d %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>% 
    slice(-c(1:2))  %>% 
    rename(tot_abroad = Total) %>% 
    mutate_at(vars(-1), as.numeric)
  
    
  clean_df[[a]] <- d
  
}

# clean the df for assessing internal migration
in_df <- list()

for (a in 1:length(clean_df)) {
  
  d <- clean_df[[a]]

in_migration <- d %>% 
  filter(current_res == "New York") %>%
    pivot_longer(cols = 2:length(d),
                 names_to = "state") %>% 
  select(!current_res)
  
  MOE <- in_migration %>%
    filter(str_detect(state, "MOE")) %>%
    rename(stat = 1,
           moe = 2)
  
  in_migration <- in_migration %>%
    filter(!str_detect(state, "MOE")) %>%
    filter(!str_detect(state, "current_res")) %>%
    cbind(MOE) %>%
    select(!stat)
  
  in_df[[a]] <- in_migration
}

s23 <- in_df[[1]]%>% 
  rename(value_23 = value,
         moe_23 = moe) %>% 
  arrange(-value_23)
s22 <- in_df[[2]] %>% 
  rename(value_22 = value,
         moe_22 = moe) %>% 
  arrange(-value_22)
s21 <- in_df[[3]] %>% 
  rename(value_21= value,
         moe_21 = moe) %>% 
  arrange(-value_21)
s19 <- in_df[[4]] %>% 
  rename(value_19 = value,
         moe_19 = moe) %>% 
  arrange(-value_19)

```

## Total in/out migration migration (including Puerto Rico)
```{r}
## testing extracting in and out 
total_in <- clean_df[[1]] %>% 
  filter(current_res %in% "New York") %>% 
  select(-matches("MOE")) %>% 
  mutate(tot_in = tot_dif_state + tot_abroad + `Puerto Rico`,
         year = 2023) %>% 
  select(year, tot_in)

total_out <- clean_df[[1]] %>% 
  select(current_res, `New York`) %>% 
  slice(-c(54:63)) %>% 
  pivot_wider(names_from = current_res,
              values_from = `New York`) %>% 
  mutate(tot_out = rowSums(select(., 2:53), na.rm = TRUE),
         year = 2023) %>% 
  select(year, tot_out)

net_migration <- total_in %>% 
  left_join(total_out, by = "year")
```

# Table 1: Total in and out migration 
```{r}
# creating loop to iterate over each year 

# clean the df for assessing internal migration
years <- seq(from = 2015, to = 2023, by = 1)
years <- setdiff(years, 2020)

# Use lapply to collect all the data frames into a list
clean_df <- lapply(years, function(year) {
  # Get the data frame for the year dynamically
  get(paste0("migration_", year))
})

net_df <- list()

for (a in 1:length(clean_df)) {
  
  d <- clean_df[[a]]
  
  total_in <- d %>% 
    filter(current_res %in% "New York") %>% 
    select(-matches("MOE")) %>% 
    mutate(tot_in = tot_dif_state + tot_abroad + `Puerto Rico`,
           year = years[a]) %>% 
    select(year, tot_pop, tot_in)
  
  total_out <- d %>% 
    select(current_res, `New York`) %>% 
    slice(-c(54:63)) %>% 
    pivot_wider(names_from = current_res,
                values_from = `New York`) %>% 
    mutate(tot_out = rowSums(select(., 2:53), na.rm = TRUE),
           year = years[a]) %>% 
    select(year, tot_out)
  
  net_migration <- total_in %>% 
    left_join(total_out, by = "year")
  
  net_df[[a]] <- net_migration
}

net_migration <- bind_rows(net_df) %>% 
  mutate(net = tot_in - tot_out,
         net_perc = ((tot_in - tot_out)/tot_pop)*100) %>% 
  arrange(year)

net_migration$percentage_change <- NA  # Initialize a column for percentage change

# Loop over rows starting from the second row (skip the first year)
for (i in 2:nrow(net_migration)) {  # Start from the second row (2022)
  # Calculate the percentage change based on the previous row's value
  net_migration$percentage_change[i] <- round(((net_migration$net[i] - net_migration$net[i-1]) / net_migration$net[i-1]) * 100, 2)
}

#colnames(net_migration) <- c("Year", "Total Population", "In", "Out", "Net Movement", "Net Percent", "Year-to-Year Change in Net Movement")

net_migration_table <- gt(net_migration) %>% 
   tab_header(
    title = "Total In and Out Migration",
    subtitle = "Net migration to and from New York Sate 2015-2023."
  ) %>% 
  cols_label(
    year = "Year",
    tot_pop = "Total Population",
    tot_in = "In",
    tot_out = "Out",
    net = "Net Movement", 
    net_perc = "Net Percent",
    percentage_change = "Year-to-Year Change in Net Movement"
  ) %>% 
  tab_source_note(
    source_note = md("**Table 1: Total migration of peolpe in and out of NYS, and the year to year change in net movement.** Negative values indicate that there was a decrease in the net number of people migrating in the current year compared to the previous year. Positive values indicate there was an increase in the net number of people migrating in the current year compared to the previous year.")
  )

write.csv(net_migration_table, here("outputs", "net_migration.csv"), row.names = FALSE)
# Save the table as an HTML file first
gtsave(net_migration_table, here("outputs", "net_migration_table.html"))

```

## Time trend Graphs

```{r}
graph1 <- net_migration %>% 
  select(year, tot_in, tot_out) %>% 
  pivot_longer(cols = 2:3,
               names_to = "flow",
               values_to = "pop") %>% 
  ggplot(aes(x = year, 
             y = pop/1000,
             color = flow)) +
  geom_line() +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("tot_in" = "#0E3B43", "tot_out" = "#B3B240"),
                     labels = c("tot_in" = "In-Migration", "tot_out" = "Out-Migration")) +  # Update legend labels
  labs(y = "Population (in 1000s)",
       x = "Year",
       color = "",
       title = str_wrap("Figure 1: Time trends for in-migration and out-migration of people in NYS between 2015 and 2023."),
       caption = str_wrap("Notes: In-Migration refers to the number of people who's residency changed to New York State from other states, territories, and foreign countries in the last year. Out-Migration refers to the number of people who's residency changed from New York State to a different state (including Puerto Rico) in the last year. 2020 migration data is not available from the Census Bureau.")) +
  scale_x_continuous(
    breaks = unique(net_migration$year),   # Explicitly set breaks for each year
    labels = as.character(unique(net_migration$year))  # Show years as labels
  ) +
  theme_bw() +
  theme(
    plot.caption = element_text(size = 10, hjust = 0, lineheight = 1.2),  # Adjust font size and line height
    plot.margin = margin(10, 10, 20, 10)  # Add margin space around the plot
  ) 

graph1

ggsave(here("outputs", "timeTrend_migration.png"), plot = graph1, width = 10, height = 7, dpi = 300)

graph2 <- net_migration %>% 
  select(year, net) %>%
  ggplot(aes(x = year, 
             y = net/1000)) +
  geom_col(fill = "#606C38") +
   labs(y = "Population (in 1000s)",
       x = "Year",
       color = "",
       title = str_wrap("Figure 2: Time trends for net migration of people in and out of NYS between 2015 and 2023."),
       caption = str_wrap("Notes: Negative values in each year indicate that every year from 2015-2023, more people are migrating out of NYS than migrating in. 2020 migration data is not available from the Census Bureau.")) +
   scale_x_continuous(
    breaks = unique(net_migration$year),   # Explicitly set breaks for each year
    labels = as.character(unique(net_migration$year))  # Show years as labels
  ) +
  theme_bw() +
  theme(
    plot.caption = element_text(size = 10, hjust = 0, lineheight = 1.2),  # Adjust font size and line height
    plot.margin = margin(10, 10, 20, 10)  # Add margin space around the plot
  )

graph2

ggsave(here("outputs", "timeTrend_net.png"), plot = graph2, width = 10, height = 10, dpi = 300)
```


# Table 2: Top 5 sources for in-migration to NY
```{r}
s19_s23 <- s23 %>% 
  left_join(s22, by = "state") %>% 
  left_join(s21, by = "state") %>% 
  left_join(s19, by = "state") %>% 
  select(-which(grepl("moe", colnames(.), ignore.case = TRUE)))

colnames(s19_s23) <- c("State", 2023, 2022, 2021, 2019)

in_migration <- s19_s23 %>% 
     slice(6:11) %>% 
  mutate(year_change = `2023`-`2022`,
         year5_change = `2023`-`2019`) %>% 
   select(State, year_change, year5_change, everything())

in_table <- gt(in_migration) %>% 
  tab_header(
    title = "In-Migration to New York State",
    subtitle = "The top five states that people are moving from to NYS 2019-2023."
  )  %>% 
  tab_spanner(
    label = "Total Number of People Moving to NYS",
    columns = vars(`2023`, `2022`, `2021`, `2019`)
  ) %>%
  cols_label(
    State = "Source",
    year_change = "1-yr Change",
    year5_change = "5-yr Change"
  ) %>% 
  tab_source_note(
    source_note = md("**Table 2: Top 5 states and foreign migration into New York State from 2019-2023.** In-Migration refers to the number of people who's residency changed to New York State from other states, territories, and foreign countries in the last year. 2020 migration data is not available from the Census Bureau.")
  )

in_table
# done - add "Total Number of People Moving to NYS"
# done - additional column with change from 2023-2022 (1yr Change) and 2023-2019 (5yr Change)

write.csv(in_table, here("outputs", "in_migration.csv"), row.names = FALSE)
# Save as HTML file
gtsave(in_table, here("outputs/in_migration.html"))
```


## Graph 3: In Migration
```{r}

state_in <- s19_s23 %>% 
     slice(6:11) %>% 
  pivot_longer(!State) %>% 
  rename(year = name, 
         pop = value) %>% 
  ggplot(aes(x = year, y = pop/1000, fill = reorder(State, pop))) +
  geom_col(position = "dodge") +
  labs(title = "Figure 3: Top Sources of Migration Into New York State from 2019 to 2023", 
       x = "Year", 
       y = "Persons (in 1000s)",
       fill = "Source",
       caption = str_wrap("Notes: In-Migration refers to the number of people who's residency changed to New York State from other states, territories, and foreign countries in the last year. 2020 migration data is not available from the Census Bureau.")) +
  viridis::scale_fill_viridis(discrete = TRUE) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 170)) +
  theme_bw() +
  theme(
    plot.caption = element_text(size = 10, hjust = 0, lineheight = 1.2),  # Adjust font size and line height
    plot.margin = margin(15, 10, 20, 10)  # Add margin space around the plot
  )

state_in
ggsave(here("outputs", "stateIn_migration.png"), plot = state_in, width = 10, height = 6, dpi = 300)
```


### Changes between 2022 and 2023
```{r}
s19_s23 %>% 
  mutate(pop_change = `2023`-`2022`) %>% 
  arrange(-abs(pop_change))
```



Create pie chart:
-   same residence in NYS
-   different residence in NYS
-   different residence outside of NYS
-   total abroad
```{r message = FALSE, warning = FALSE}
source_df <- s19_s23 %>% 
  pivot_longer(!State) %>% 
  rename(Year = name,
         pop = value) %>% 
  slice(1:20) %>% 
  #filter(!(State %in% c("tot_pop", "tot_same_res"))) %>%
  pivot_wider(names_from = State, values_from = pop) %>% 
  select(Year, `Total Population` = tot_pop, 
         `Same Residency` = tot_same_res,
         `Within State Migration` = dif_res, 
         `Migration from Out of State` = tot_dif_state, 
         `Migration from Abroad` = 	tot_abroad) %>% 
  mutate(Year = as.numeric(Year),
         tot = `Same Residency` + `Within State Migration` + `Migration from Out of State` + `Migration from Abroad`)

# significance tests
lm(`Within State Migration` ~ Year, data = source_df) %>% 
  summary()

tableSource <- source_df %>% 
  gt() %>% 
  tab_header(
    title = "Sources of Migration Into New York State",
    subtitle = "Within-State and Out-of-State migration to NYS 2019-2023."
  )

# Save as HTML file
gtsave(tableSource, here("outputs/sourceIn_migration.html"))

ss_pie <- s19_s23 %>% 
  pivot_longer(!State) %>% 
  rename(Year = name,
         pop = value) %>% 
  slice(1:20) %>% 
  #filter(!(State %in% c("tot_pop", "tot_same_res"))) %>%
  pivot_wider(names_from = State, values_from = pop) %>% 
  mutate(perc_same = (tot_same_res/tot_pop)*100,
         perc_dif_res = (dif_res/tot_pop)*100,
         perc_dif_state = (tot_dif_state/tot_pop)*100,
         perc_abroad = (tot_abroad/tot_pop)*100) %>% 
  select(Year, `Same Residency` = perc_same, 
         `Within State Migration` = perc_dif_res, 
         `Migration from Out of State` = perc_dif_state, 
         `Migration from Abroad` = perc_abroad) %>% 
  pivot_longer(cols = 2:5,
               names_to = "category",
               values_to = "percent")

pie23 <- ss_pie %>%   
filter(Year %in% 2023) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2023",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
   # Set the plot background to white
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie23.png"), plot = pie23, width = 8, height = 6, dpi = 300)

pie22 <- ss_pie %>%   
filter(Year %in% 2022) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2022",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie22.png"), plot = pie22, width = 8, height = 6, dpi = 300)

pie21 <- ss_pie %>%   
filter(Year %in% 2021) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2021",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie21.png"), plot = pie21, width = 8, height = 6, dpi = 300)


pie19 <- ss_pie %>%   
filter(Year %in% 2019) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2019",
       fill = "") +
   scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text


# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie19.png"), plot = pie19, width = 8, height = 6, dpi = 300)

```



### Explore trends for state to state migration in 2023

The top 5 states migrating to NY in 2023: 
-   Massachusetts
-   Florida
-   Pennsylvania
-   California
-   New Jersey

```{r}
# Example ggplot with ordering by value from highest to lowest
ggplot(s19_s23 %>% 
         filter(state %in% c("Massachusetts", "Florida", "Pennsylvania", "California", "New Jersey")), 
       aes(x = reorder(state, value), y = value / 1000, fill = year)) +
  geom_col(position = "dodge") +
  labs(title = "Figure 3: Top 5 States Migrating to New York 2019-2023", 
       x = "State", 
       y = "Persons (in 1000 people)") +
  theme_minimal()

ggplot(s19_s23 %>% 
         filter(state %in% c("Massachusetts", "Florida", "Pennsylvania", "California", "New Jersey")), 
       aes(x = year, y = reorder(value / 1000, value), fill = state)) +
  geom_col(position = "dodge") +
  labs(title = "Top 5 States Migrating to New York 2019-2023", 
       x = "Year", 
       y = "Persons (per 1000 people)") +
  theme_minimal()
  
```

### Explore trends for foreign migration to NY

Decrease from 2019 to 2021, probably from COVID. Steady increase since then.

```{r}
s19_s23 %>% 
  filter(state %in% c("Foreign Country")) %>% 
  ggplot(aes(x = year, y = value/1000)) +
  geom_col(position = "dodge") +
  labs(title = "Abroad Migration to New York 2019-2023", 
       x = "Year", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

```


### Where are New Yorkers moving to?

Out of NY to other states
-   `new_prop` is a measure of the percent of the total population migrating to their current residence in 2023, that are from NYS.


# Table 3: Top 5 destination for out-migration from NY

```{r message = FALSE, warning = FALSE}
ss19 <- clean_df[[4]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2019)

ss21 <- clean_df[[3]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2021)

ss22 <- clean_df[[2]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2022)

ss23 <- clean_df[[1]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = `MOE (±)_75`) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2023)

ss19_23 <- rbind(ss19, ss21, ss22, ss23) %>% 
  mutate(perc = round((from_ny/tot_dif_state)*100, 2))

names <- c("Current Residence", "Total Population", "Total In-Migration to State", "In-Migration From NYS", "MOE", "Year", "% From NYS")

colnames(ss19_23) <- names

out_table <- ss19_23 %>% 
  select(!c(Year,MOE, `Total Population`)) 

gt(out_table) %>% 
  tab_header(
    title = "Migration Out of New York State",
    subtitle = "The top five states that people from NYS are migrating to 2019-2023."
  ) %>%
  tab_source_note(
    source_note = "% From NYS: Refers to the percent of total in migration coming from NYS."
  ) %>%
  tab_row_group(
    group = md("**2019**"),
    rows = 1:5
  ) %>% 
  tab_row_group(
    group = "2021",
    rows = 6:10
  ) %>% 
  tab_row_group(
    group = "2022",
    rows = 11:15
  ) %>% 
  tab_row_group(
    group = "2023",
    rows = 16:20
  ) %>% 
  cols_label(
    `Current Residence` = "Destination"
  ) %>% 
  tab_source_note(
    source_note = md("**Table 3: Top 5 desitnations for out-migration from New York State 2019-2023.** For each year, the top source was from foreign countries. The top 5 states migrating to NYS were New Jersey, California, Pennsylvania, Florida, Massachusetts.")
  )

write.csv(out_table, here("outputs", "out_migration.csv"), row.names = FALSE)
# Save as HTML file
# Save as HTML file
gtsave(table2, here("outputs/out_migration.html"))

```

## Graph 4
```{r }
state_out <- ss19_23 %>% 
  select(-c(MOE, `Total Population`)) %>% 
  rename(Destination = `Current Residence`) %>% 
  ggplot(aes(x = Year, y = `In-Migration From NYS`/1000, fill = reorder(Destination, `In-Migration From NYS`)))+
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(round(`% From NYS`, 0), "%")),  # Add labels above the bars
            position = position_dodge(width = 0.8), vjust = -0.5, size = 3) +
  labs(title = "Figure 4: Top Destinations of Out-Migration From New York State 2019 to 2023", 
       x = "Year", 
       y = "Persons (in 1000s)",
       fill = "",
       caption = str_wrap("Notes: 2020 migration data is not available from the Census Bureau. 2020 migration data is not available from the Census Bureau. The percentages show the contribution of total migration to that state that are from New York State. NYS consistently makes up the most of the migration to Connecticut and New Jersey.")) +
  paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto") +
  scale_x_continuous(breaks = c(2019, 2021, 2022, 2023)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 100)) +
  theme_bw() +
  theme(
    plot.caption = element_text(size = 10, hjust = 0, lineheight = 1.2),  # Adjust font size and line height
    plot.margin = margin(25, 15, 25, 15),  # Further increase top margin to provide room for labels
    plot.title = element_text(margin = margin(b = 30))
  )

state_out
ggsave(here("outputs", "stateOut_migration.png"), plot = state_out, width = 10, height = 6, dpi = 300)

```



```{r message = FALSE, warning = FALSE}
out23 <- clean_df[[1]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  rename(MOE_75 = 4) %>% 
  mutate(year = 2023)

out22 <- clean_df[[2]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2022)

out21 <- clean_df[[3]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2021)

out19 <- clean_df[[4]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2019)

out <- rbind(out19, out21, out22, out23)
```


```{r}

ggplot(out, aes(x = reorder(year, ny_value), y = ny_value, fill = current_res)) +
  geom_col(position = "dodge")

ggplot(out23 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2023", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

ggplot(out22 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2022", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

ggplot(out21 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2021", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()


ggplot(out19 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2019", 
       x = "State", 
       y = "Persons (in 1000s)") +
  theme_minimal()
```

# State of residence by birthplace

### Read in the data files and clean them
```{r}

# Directory where the Excel files are stored
file_dir <- here("data", "state_of_residence_by_birthplace")

# List all Excel files in the directory
files <- list.files(file_dir, pattern = "*.xlsx", full.names = TRUE)

# Define a function to process each file
process_file <- function(file_path) {
  # Extract year from the file name (assuming year is in the filename)
  year <- sub(".*([0-9]{4}).xlsx", "\\1", basename(file_path))
  
  # Read and process the data
  df <- read_xlsx(file_path, skip = 4) %>%
    rename(current_res = 1,
           tot_pop = 2) %>%  # different state as 1 year ago
    select(-3)
  
  # Create the column names from the state row
  for (i in seq(3, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(df[1, i])  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the moe row
  for (i in seq(4, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(df[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  df <- df %>%
    slice(-c(1:3)) %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>% 
    slice(-c(54:dim(df[1])))
  
  # Dynamically assign the processed dataframe to a variable named sb_<year>
  assign(paste0("sb_", year), df, envir = .GlobalEnv)
}

# Apply the function to all files in the folder
lapply(files, process_file)

```

## Where are people in NYS from?
```{r}
sb_2023 %>% 
  filter(current_res == "New York") %>% 
  pivot_longer(cols = 2:118,
               names_to = "state",
               values_to = "pop") %>% 
  filter(!str_detect(state, "MOE")) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  arrange(-pop) %>% 
  mutate(perc = round((pop/19571216)*100, 2))
```

```{r}
list_df <- list(sb_2023, sb_2022, sb_2021, sb_2019)
# Initialize an empty dataframe to hold the final combined result
final_df <- data.frame()
years <- c(2023, 2022, 2021, 2019)

for (a in 1:length(list_df)) {
  
  d <- list_df[[a]]
  year <- gsub("sb_", "", names(list_df)[a])  # Extract year from the list name (sb_2023 -> 2023)
  
  d <- d %>% 
    filter(current_res == "New York") %>% 
    pivot_longer(cols = 2:118,
                 names_to = "state",
                 values_to = "pop") %>% 
    filter(!str_detect(state, "MOE")) %>% 
    mutate(pop = as.numeric(pop)) %>% 
    arrange(-pop) %>% 
    mutate(year = years[a])
  
  # Bind the current data frame (d) into the final_df
  final_df <- bind_rows(final_df, d)
  
}

birth_df <- final_df %>%
  select(!current_res) %>% 
  pivot_wider(names_from = state,
              values_from = pop) 

birth_df %>% 
  pivot_longer(cols = 3:60,
               names_to = "state",
               values_to = "pop") %>% 
  mutate(perc = round((pop/tot_pop)*100, 2)) %>% 
  #group_by(year, state) %>% 
  arrange(-perc) %>% 
  select(!tot_pop)
```

## What percentage of people are born in the state that they are currently residing in?

tot_pop = total population of state
pop = population where birthplace is the same as that state
```{r }
s <- list_df[[1]]$current_res

states_names <- colnames(birth_df)

states_names <- intersect(s, states_names)

#states_names <-states_names[5:60]


filtered_states_data <- list()

for (place in states_names) {
  
  print(place)
  # Filter and process the data for each state
  filtered_data <- list_df[[1]] %>%
    filter(current_res == place) %>%  # Filter for the current state
    pivot_longer(cols = 2:118,  # Adjust column range as needed
                 names_to = "state", 
                 values_to = "pop") %>%
    filter(!str_detect(state, "MOE")) %>%  # Remove columns with "MOE" in the name
    mutate(pop = as.numeric(pop)) %>%  # Convert population to numeric
    arrange(-pop) %>%  # Sort by descending population
    pivot_wider(names_from = state,
                values_from = pop) %>% 
    pivot_longer(cols = 3:60,
                 names_to = "state",
                 values_to = "pop") %>% 
    mutate(perc = round((pop/tot_pop)*100, 2)) %>% 
    filter(state %in% place)
  
  # Store the filtered data in the list using the state name as the list key
  filtered_states_data[[place]] <- filtered_data
}

local_df <- bind_rows(filtered_states_data)
```


