---
title: "Investigating Migration Flows in and out of NYS"



---

```{r load libraries, message = FALSE}
library(tidyverse)
library(readxl)
library(here)
library(gt)
```

# Read in Data

## State to state migration Data
```{r message = FALSE, warning = FALSE}
ss_23 <- read_xlsx(here("data", "state_to_state_migration/2023.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         ) %>% 
  slice(-1)

# I had to clean this data before I imported it because it had lots of messy tabls that did not download well
ss_22 <- read_xlsx(here("data", "state_to_state_migration/2022.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )

# I had to clean this data before I imported it because it had lots of messy tabls that did not download well
ss_21 <- read_xlsx(here("data", "state_to_state_migration/2021.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )
ss_19 <- read_xlsx(here("data", "state_to_state_migration/2019.xlsx"), skip = 4) %>% 
  rename(current_res = 1,
         tot_pop = 2,
         tot_same_res = 4, # same residence as 1 year ago
         dif_res = 6, # different residence / same state as 1 year ago
         tot_dif_state = 8 # different state as 1 year ago
         )


## to remove rows that contain all NAs
#filter(rowSums(is.na(ss_22)) != ncol(ss_22))
```


## State to state migration

2023:
Net migration 
Into NY from other states
Into NY from Abroad

### Clean data
```{r message = FALSE, warning = FALSE}
list_df <- list(ss_23, ss_22, ss_21, ss_19)
clean_df <- list()

for (a in 1:length(list_df)) {
  
  d <- list_df[[a]]
  
  # Create the column names from the state row
  for (i in seq(10, ncol(d), by = 2)) {
    if (i <= ncol(d)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(d[1, i])  # Get the value from row 1 of the ith column
      colnames(d)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the state row
  for (i in seq(3, ncol(d), by = 2)) {
    if (i <= ncol(d)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(d[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(d)[i] <- col_name  # Set column name to that value
    }
  }
  
  d <- d %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>% 
    slice(-c(1:2))  %>% 
    rename(tot_abroad = Total)
  
    
  clean_df[[a]] <- d
  
}

# clean the df for assessing internal migration
in_df <- list()

for (a in 1:length(clean_df)) {
  
  d <- clean_df[[a]]

in_migration <- d %>% 
  filter(current_res == "New York") %>%
    pivot_longer(cols = 1:length(d),
                 names_to = "state")
  
  MOE <- in_migration %>%
    filter(str_detect(state, "MOE")) %>%
    rename(stat = 1,
           moe = 2)
  
  in_migration <- in_migration %>%
    filter(!str_detect(state, "MOE")) %>%
    filter(!str_detect(state, "current_res")) %>%
    cbind(MOE) %>%
    select(!stat) %>%
    mutate(value = as.numeric(value),
           moe = as.numeric(moe))
  
  in_df[[a]] <- in_migration
}

s23 <- in_df[[1]]%>% 
  rename(value_23 = value,
         moe_23 = moe) %>% 
  arrange(-value_23) %>% 
  slice(1:11)
s22 <- in_df[[2]] %>% 
  rename(value_22 = value,
         moe_22 = moe) %>% 
  arrange(-value_22) %>% 
  slice(1:11)
s21 <- in_df[[3]] %>% 
  rename(value_21= value,
         moe_21 = moe) %>% 
  arrange(-value_21) %>% 
  slice(1:11)
s19 <- in_df[[4]] %>% 
  rename(value_19 = value,
         moe_19 = moe) %>% 
  arrange(-value_19) %>% 
  slice(1:11)


s19_s23 <- s23 %>% 
  left_join(s22, by = "state") %>% 
  left_join(s21, by = "state") %>% 
  left_join(s19, by = "state") %>% 
  select(-which(grepl("moe", colnames(.), ignore.case = TRUE)))

colnames(s19_s23) <- c("State", 2023, 2022, 2021, 2019)

table1 <- gt(s19_s23 %>% 
     slice(6:11)) %>% 
  tab_header(
    title = "Migration Into New York State",
    subtitle = "The top five states that people are moving from to NYS 2019-2023."
  )

# Save as HTML file
gtsave(table1, here("outputs/in_migration.html"))
```

Create pie chart:
-   same residence in NYS
-   different residence in NYS
-   different residence outside of NYS
-   total abroad
```{r message = FALSE, warning = FALSE}
tableSource <- s19_s23 %>% 
  pivot_longer(!State) %>% 
  rename(Year = name,
         pop = value) %>% 
  slice(1:20) %>% 
  #filter(!(State %in% c("tot_pop", "tot_same_res"))) %>%
  pivot_wider(names_from = State, values_from = pop) %>% 
  select(Year, `Total Population` = tot_pop, 
         `Same Residency` = tot_same_res,
         `Within State Migration` = dif_res, 
         `Migration from Out of State` = tot_dif_state, 
         `Migration from Abroad` = 	tot_abroad) %>% 
  gt() %>% 
  tab_header(
    title = "Sources of Migration Into New York State",
    subtitle = "Within-State and Out-of-State migration to NYS 2019-2023."
  )

# Save as HTML file
gtsave(tableSource, here("outputs/sourceIn_migration.html"))

ss_pie <- s19_s23 %>% 
  pivot_longer(!State) %>% 
  rename(Year = name,
         pop = value) %>% 
  slice(1:20) %>% 
  #filter(!(State %in% c("tot_pop", "tot_same_res"))) %>%
  pivot_wider(names_from = State, values_from = pop) %>% 
  mutate(perc_same = (tot_same_res/tot_pop)*100,
         perc_dif_res = (dif_res/tot_pop)*100,
         perc_dif_state = (tot_dif_state/tot_pop)*100,
         perc_abroad = (tot_abroad/tot_pop)*100) %>% 
  select(Year, `Same Residency` = perc_same, 
         `Within State Migration` = perc_dif_res, 
         `Migration from Out of State` = perc_dif_state, 
         `Migration from Abroad` = perc_abroad) %>% 
  pivot_longer(cols = 2:5,
               names_to = "category",
               values_to = "percent")

pie23 <- ss_pie %>%   
filter(Year %in% 2023) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2023",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
   # Set the plot background to white
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie23.png"), plot = pie23, width = 8, height = 6, dpi = 300)

pie22 <- ss_pie %>%   
filter(Year %in% 2022) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2022",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie22.png"), plot = pie22, width = 8, height = 6, dpi = 300)

pie21 <- ss_pie %>%   
filter(Year %in% 2021) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2021",
       fill = "") +
  scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text

# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie21.png"), plot = pie21, width = 8, height = 6, dpi = 300)


pie19 <- ss_pie %>%   
filter(Year %in% 2019) %>% 
  ggplot(aes(x = "", y = percent, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  theme_void() +
  labs(title = "Sources of Migration to NYS 2019",
       fill = "") +
   scale_fill_brewer(palette = "Set1") +
  theme(
    plot.background = element_rect(fill = "white")  # White background
  ) +
  ggrepel::geom_text_repel(aes(label = paste0(round(percent, 1), "%")), 
                  size = 4, 
                   position = position_stack(vjust = 0.4),
                  #box.padding = 0.5,  # Adjust the space around the labels
                  color = "white",    # Label color
                  show.legend = FALSE)  # Hide legend for text


# Save the plot to a file in the "output" folder
ggsave(here("outputs/source_pie19.png"), plot = pie19, width = 8, height = 6, dpi = 300)

```



### Explore trends for state to state migration in 2023

The top 5 states migrating to NY in 2023: 
-   Massachusetts
-   Florida
-   Pennsylvania
-   California
-   New Jersey

```{r}
# Example ggplot with ordering by value from highest to lowest
ggplot(s19_s23 %>% 
         filter(state %in% c("Massachusetts", "Florida", "Pennsylvania", "California", "New Jersey")), 
       aes(x = reorder(state, value), y = value / 1000, fill = year)) +
  geom_col(position = "dodge") +
  labs(title = "Top 5 States Migrating to New York 2019-2023", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

ggplot(s19_s23 %>% 
         filter(state %in% c("Massachusetts", "Florida", "Pennsylvania", "California", "New Jersey")), 
       aes(x = year, y = reorder(value / 1000, value), fill = state)) +
  geom_col(position = "dodge") +
  labs(title = "Top 5 States Migrating to New York 2019-2023", 
       x = "Year", 
       y = "Persons (per 1000 people)") +
  theme_minimal()
  
```

### Explore trends for foreign migration to NY

Decrease from 2019 to 2021, probably from COVID. Steady increase since then.

```{r}
s19_s23 %>% 
  filter(state %in% c("Foreign Country")) %>% 
  ggplot(aes(x = year, y = value/1000)) +
  geom_col(position = "dodge") +
  labs(title = "Abroad Migration to New York 2019-2023", 
       x = "Year", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

```


### Where are New Yorkers moving to?

Out of NY to other states
-   `new_prop` is a measure of the percent of the total population migrating to their current residence in 2023, that are from NYS.

# Tables: Migration Out of New York State

```{r message = FALSE, warning = FALSE}
ss19 <- clean_df[[4]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2019)

ss21 <- clean_df[[3]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2021)

ss22 <- clean_df[[2]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = MOE_75) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2022)

ss23 <- clean_df[[1]] %>% 
  select(current_res, tot_pop, tot_dif_state, from_ny = `New York`, moe = `MOE (±)_75`) %>% 
  slice(-c(54:62)) %>% 
  mutate(tot_dif_state = as.numeric(tot_dif_state),
         tot_pop = as.numeric(tot_pop),
         from_ny = as.numeric(from_ny),
         moe = as.numeric(moe)) %>% 
  arrange(-from_ny) %>% 
  slice(2:6) %>% 
  mutate(year = 2023)

ss19_23 <- rbind(ss19, ss21, ss22, ss23) %>% 
  mutate(perc = round((from_ny/tot_dif_state)*100, 2))

names <- c("Current Residence", "Total Population", "Total In Migration", "Migration From NYS", "MOE", "Year", "% From NYS")

colnames(ss19_23) <- names

ss_table <- ss19_23 %>% 
  select(!c(Year,MOE)) 

table2 <-gt(ss_table) %>% 
  tab_header(
    title = "Migration Out of New York State",
    subtitle = "The top five states that people from NYS are migrating to 2019-2023."
  ) %>%
  tab_source_note(
    source_note = "% From NYS: Refers to the percent of total in migration coming from NYS."
  ) %>%
  tab_row_group(
    group = "2019",
    rows = 1:5
  ) %>% 
  tab_row_group(
    group = "2021",
    rows = 6:10
  ) %>% 
  tab_row_group(
    group = "2022",
    rows = 11:15
  ) %>% 
  tab_row_group(
    group = "2023",
    rows = 16:20
  )

# Save as HTML file
gtsave(table2, here("outputs/out_migration.html"))

```


```{r message = FALSE, warning = FALSE}
out23 <- clean_df[[1]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  rename(MOE_75 = 4) %>% 
  mutate(year = 2023)

out22 <- clean_df[[2]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2022)

out21 <- clean_df[[3]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2021)

out19 <- clean_df[[4]] %>% 
  select(1:9, "New York", 75) %>% 
  mutate(across(-1, as.numeric)) %>% 
  slice(-c(54:63)) %>% 
  rename(ny_value = 10) %>% 
  # the % of the people coming from NY that make up the migration to these states
  mutate(new_prop = (tot_dif_state/tot_pop)*100) %>% 
  arrange(-ny_value) %>%
  head(11) %>% 
  slice(-1) %>%  # remove total pop
  select(current_res, tot_pop, 10:12) %>% 
  mutate(year = 2019)

out <- rbind(out19, out21, out22, out23)
```


```{r}

ggplot(out, aes(x = reorder(year, ny_value), y = ny_value, fill = current_res)) +
  geom_col(position = "dodge")

ggplot(out23 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2023", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

ggplot(out22 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2022", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()

ggplot(out21 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2021", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()


ggplot(out19 %>% 
         slice(-1) %>% 
         head(9), 
       aes(x = reorder(current_res, ny_value), y = ny_value/1000)) +
  geom_col() +
  labs(title = "Where NYS residents are moving to 2019", 
       x = "State", 
       y = "Persons (per 1000 people)") +
  theme_minimal()
```

## State of residence by birthplace

### Read in the data files and clean them
```{r}

# Directory where the Excel files are stored
file_dir <- here("data", "state_of_residence_by_birthplace")

# List all Excel files in the directory
files <- list.files(file_dir, pattern = "*.xlsx", full.names = TRUE)

# Define a function to process each file
process_file <- function(file_path) {
  # Extract year from the file name (assuming year is in the filename)
  year <- sub(".*([0-9]{4}).xlsx", "\\1", basename(file_path))
  
  # Read and process the data
  df <- read_xlsx(file_path, skip = 4) %>%
    rename(current_res = 1,
           tot_pop = 2) %>%  # different state as 1 year ago
    select(-3)
  
  # Create the column names from the state row
  for (i in seq(3, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- as.character(df[1, i])  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  # Create the column names from the moe row
  for (i in seq(4, ncol(df), by = 2)) {
    if (i <= ncol(df)) {  # Ensure we don't go beyond the number of columns
      col_name <- paste0(as.character(df[2, i]), "_", i)  # Get the value from row 1 of the ith column
      colnames(df)[i] <- col_name  # Set column name to that value
    }
  }
  
  df <- df %>%
    slice(-c(1:3)) %>%
    filter(rowSums(is.na(.)) != ncol(.)) %>% 
    slice(-c(54:dim(df[1])))
  
  # Dynamically assign the processed dataframe to a variable named sb_<year>
  assign(paste0("sb_", year), df, envir = .GlobalEnv)
}

# Apply the function to all files in the folder
lapply(files, process_file)

```

## Where are people in NYS from?
```{r}
sb_2023 %>% 
  filter(current_res == "New York") %>% 
  pivot_longer(cols = 2:118,
               names_to = "state",
               values_to = "pop") %>% 
  filter(!str_detect(state, "MOE")) %>% 
  mutate(pop = as.numeric(pop)) %>% 
  arrange(-pop) %>% 
  mutate(perc = round((pop/19571216)*100, 2))
```

```{r}
list_df <- list(sb_2023, sb_2022, sb_2021, sb_2019)
# Initialize an empty dataframe to hold the final combined result
final_df <- data.frame()
years <- c(2023, 2022, 2021, 2019)

for (a in 1:length(list_df)) {
  
  d <- list_df[[a]]
  year <- gsub("sb_", "", names(list_df)[a])  # Extract year from the list name (sb_2023 -> 2023)
  
  d <- d %>% 
    filter(current_res == "New York") %>% 
    pivot_longer(cols = 2:118,
                 names_to = "state",
                 values_to = "pop") %>% 
    filter(!str_detect(state, "MOE")) %>% 
    mutate(pop = as.numeric(pop)) %>% 
    arrange(-pop) %>% 
    mutate(year = years[a])
  
  # Bind the current data frame (d) into the final_df
  final_df <- bind_rows(final_df, d)
  
}

birth_df <- final_df %>%
  select(!current_res) %>% 
  pivot_wider(names_from = state,
              values_from = pop) 

birth_df %>% 
  pivot_longer(cols = 3:60,
               names_to = "state",
               values_to = "pop") %>% 
  mutate(perc = round((pop/tot_pop)*100, 2)) %>% 
  #group_by(year, state) %>% 
  arrange(-perc)
  pivot_wider(names_from = year,
              values_from = c(pop, perc))
```


```{r}


states_names <- colnames(birth_df)

states_names <-states_names[5:60]

filtered_states_data <- list()

for (place in states_names) {
  # Filter and process the data for each state
  filtered_data <- list_df[[1]] %>%
    filter(current_res == place) %>%  # Filter for the current state
    pivot_longer(cols = 2:118,  # Adjust column range as needed
                 names_to = "state", 
                 values_to = "pop") %>%
    filter(!str_detect(state, "MOE")) %>%  # Remove columns with "MOE" in the name
    mutate(pop = as.numeric(pop)) %>%  # Convert population to numeric
    arrange(-pop) %>%  # Sort by descending population
    pivot_wider(names_from = state,
                values_from = pop) %>% 
    pivot_longer(cols = 3:60,
                 names_to = "state",
                 values_to = "pop") %>% 
    mutate(perc = round((pop/tot_pop)*100, 2)) %>% 
    filter(state %in% place)
  
  # Store the filtered data in the list using the state name as the list key
  filtered_states_data[[place]] <- filtered_data
}
```


